# (PART) Rotations {-}

# Rotation A: e-Cigarette Liquid

##### **Objective** {-}

Read the Memorandum on the following page to understand the scenario for this lab and determine your objectives.

##### **To Turn In** {-}

- Separate lab notebooks for each technique (e.g. one for HPLC, one for FT-IR, etc.) as ZIP folders containing your notebook and data.  Each student will submit a notebook individually roughly every three weeks.
- A technical report detailing the entire project, to be turned in in stages (one per group).
- Consult the [schedule](/archive/chem370-f2020/schedule) to keep track of due dates for notebooks and report stages.  
- Refer to the materials on the [writing page](/archive/chem370-f2020/writing) for templates, formatting guidelines, content expectations, and grading rubrics.
- An analysis worksheet for each instrument (linked under the Data Processing sections for each instrument).

##### **Student Learning Outcomes** | *By the end of this lab, students will:*    {-}

- Utilize chemical instrumentation for qualitative analysis of complex samples (e.g. GC-MS, Flame AA to ID sample contents).  
- Apply instrument calibration techniques to conduct quantitative analysis on complex samples (e.g. FT-IR, HPLC, and Flame AA).   
- Assess data validity using quality control and blank samples.  
- Utilize R and RStudio to perform data analysis in an electronic lab notebook (including chemometric analysis).  
- Participate effectively in group lab work and writing activities.  
- Compile results from multiple analyses into a unified report that clearly conveys conclusion(s) to the question(s) posed at project start.  

\newpage

## Memorandum {-}

\noindent
***Western Analytics, Inc.***

\noindent  
**To:** Aerosol Exposure Division  
\noindent  
**From:** Al Fischer, PhD    
\noindent  
**Subject:** e-Cigarette Liquid Analysis   

### Western Analysts: {-}

The Centers for Disease Control (CDC) reports that an outbreak of e-cigarette, or vaping, associated lung injury (EVALI) began in June 2019.  It continues today, and will continue until a culprit is found.  As of December 17, 2019 a total of **2,506 EVALI hospitalizations** from 50 states, the District of Columbia, and two U.S. territories (Puerto Rico and U.S. Virgin Islands), and **54 EVALI deaths** from 27 US states have been reported to CDC. Studies have found vitamin E acetate, plant and vegetable oils, petroleum distillates, and terpenes in the bronchoalveolar lavage fluid of many affected individuals. [1]  The figure below shows some common chemical exposures from e-cigarettes.

![Common exposure pathways related to e-cigarettes.  Public domain image from [CDC Stacks](https://stacks.cdc.gov/gsearch?collection=&terms=CS292347-A).](../assets/img/e-cigarette-cdc.jpg)

The North Carolina Department of Health and Human Services (NCDHHS) is seeking our help in finding the cause of the outbreak in NC. (See [this WLOS story](https://wlos.com/news/local/nc-health-officials-report-3-cases-of-severe-lung-disease-after-vaping) for additional information.) We are asked to screen e-cigarette liquid for toxicants.

##### *Your job is to screen the products for:*  

  1. Flavoring compounds using GC-MS  
  2. Nicotine concentration using HPLC  
  3. Zinc concentration using flame AA (before and after vaping)  
  4. General adulteration using FTIR and the Hit Quality Index method in [Rodriguez et al 2011](https://www.researchgate.net/profile/John_Clarkson6/post/What_are_some_good_reference_bands_for_Raman_spectroscopy/attachment/59d6244c79197b8077982a27/AS%3A312109580128257%401451424130311/download/Standardization+of+Raman+spectra+for+transfer+of+spectral+libraries+across+different+instruments.pdf).   

Please turn in a technical report detailing the findings of your analyses.  Your report should link together the results from all your analyses, use complementary results to bolster your conclusions, and discuss what implications the findings from each analysis have on human health.

##### **Sample Information:** {-}

NCDHHS asked EVALI patients which types of of vaping liquids they used.  NCDHHS acquired these products and sent them to Western Analytics for characterization.  Aliquots of two samples will be randomly assigned to your workgroup for analysis.  Some samples are regulated off-the-shelf e-cigarette liquid and some may be home-made or "black-market" samples.  You may find problems with one, both, or none of your samples.  You will also be given a toxicant-free reference sample.  Although samples are over-the-counter products, they are all presumed toxic; you should avoid ingesting and/or contacting them with your skin.

##### **Record Keeping:**  {-}

Due to the high-impact nature of this work, you must diligently keep track of all sample information (e.g. samples names) and data.  You must also follow our standard quality-control and data-validation practices to ensure your results are accurate.  Likewise, all results must meet the precision guidelines described in the procedure for each instrument.

##### *Keep in mind:*   {-}

- There are no additional samples available -- plan ahead to make sure you don't use up all your sample!  
- The lot number for each sample (provided on the bag/vial) must be explicitly tracked in your lab notebook and final report.  
- You must provide me with your data files and lab notebooks (in addition to your report) for validation.  
- You must get certified on each instrument prior to using it.  

This is of utmost importance.  The sooner the culprit(s) is (are) found, the sooner we can help get dangerous products off the shelf, and the more lives we can save!

Please visit my office or email me with any questions.

### References {-}

[1] [CDC: Outbreak of Lung Injury Associated with the Use of E-Cigarette, or Vaping, Products](https://www.cdc.gov/tobacco/basic_information/e-cigarettes/severe-lung-disease.html)
[2] [Journal Article](https://academic.oup.com/ntr/article/17/2/134/1258864)

\newpage

## GC-MS

Gas chromatography is a technique used to separate mixtures of gases.  The gases are injected into a column, which (in our case) is a long, hollow tube with a non-polar liquid coating on the inside.  They are pushed through the column with a carrier gas, He in our instrument.  The column is initially held at low temperature (a little above room temperature) and ramped at a defined rate up to 250 degrees C.  The combination of the temperature ramp and the non-polar column allows compounds to be separated based on boiling point and polarity.  GC can be paired with many different techniques, and is often used with electron-ionization mass spectrometry.  This type of mass spectrometer breaks molecules apart into fragments, sorts the fragments by mass, and counts the number of fragments of each mass.  The specific spectrum (or distribution of fragments) is unique for every chemical compound and is based on the stability of each bond in the molecule.  In this way, the identity of a compound can usually be deduced from a mass spectrum alone, although techniques such as nuclear magnetic resonance spectroscopy and Fourier-transform infrared spectroscopy can greatly aid in identification.

You will conduct liquid phase microextraction to analyze the volatile constituents of e-cigarette liquid.  You should identify the major flavor constituents and comment on whether or not they may be harmful at the approximate levels present.

The general order for GC-MS sampling is: (1) run solvent blanks on the GC-MS until the instrument is clean,  (2) prepare a single sample using liquid-phase microextraction, (3) run the sample, and (4) repeat 2-4 as necessary.  You should obtain a clean blank before preparing your samples.  For GC-MS, a passing blank will have a baseline less than 50,000 counts that is stable and peak-free.  You may see an increase in the baseline at temperature above 250$^{\circ}$C; this is normal and still considered passing.  If the blank is dirty you will not be able to tell what is from your sample and what is contamination!  

This is a qualitative analysis.

### Safety {-}

- This lab uses small amounts of methanol ([SDS](http://www.sciencelab.com/msds.php?msdsId=9927227)).  Methanol can cause blindness and absorbs through the skin.  Wear gloves and goggles when working with methanol, and use it in a hood.  
- Microliter syringes are very sharp -- avoid poking yourself!
- Although samples are over-the-counter products, you should avoid ingesting and/or contacting them with your skin.

### Instrumentation {#gcms-instrumentation}

1. Run a solvent blank on the GC-MS using the `CHEM370_Vape.M` method.  Use a microliter syringe to inject your sample and consult the GC-MS SOP.

  ```{marginfigure}
  ***Checkpoint:***

  A solvent blank in this case is 1 $\mu$L methanol and nothing else.

  Your blank should have a baseline less than 50,000 counts that is stable and free of peaks.  If you see peaks, run another blank until you don't!
  ```

2. If your blank is clean, prepare a sample as per below and run on the GC-MS using the `CHEM370_Vape.M` method.  Consult the GC-MS SOP for more information.

### Sample Preparation {#gcms-sample-prep}

1. Place 5 drops of the e-cigarette liquid in a 2-mL GC autosampler vial.
1. Clean a microliter syringe with the syringe cleaning apparatus.
1. Load the syringe with 2 $\mu$L of methanol.
1. Place the syringe in an arm clamp/ring stand and push the GC autosampler vial onto the tip of the syringe.  You want the tip of the syringe to be about 1.5 mm from the top of the sample.

    > Do NOT let the tip touch the sample or the vial!

1. *Gently* push the plunger of the syringe down all the way so that you have a small drop of methanol on the end of the syringe.  Do not let the drop touch anything!

    > The syringe plunger does not go all the way down to the glass body!  Do not force it or you will ruin the syringe!

1. Expose the drop of solvent for 30 seconds and then pull the plunger back up to pull the methanol back into the syringe.  Do not bump the syringe, sample, or lab bench during this time.
1. Inject the sample into the GC immediately.
1. Clean the syringe and repeat as-needed for each sample.

### Data Processing {#gcms-data-processing-vape}

The following videos walk you through the data processing portion of this lab.

<iframe src="https://wcu.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=7d2b5841-5888-453d-9173-ac6300e32d05&autoplay=false&offerviewer=true&showtitle=true&showbrand=false&start=0&interactivity=all" height="405" width="720" frameBorder = "0" style="border: 0px solid #464646; display: block; margin: auto;" allowfullscreen allow="autoplay"></iframe>


<iframe src="https://wcu.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=d8e961a2-7075-437e-815e-ac6300e32d6d&autoplay=false&offerviewer=true&showtitle=true&showbrand=false&start=0&interactivity=all" height="405" width="720" frameBorder = "0" style="border: 0px solid #464646; display: block; margin: auto;" allowfullscreen allow="autoplay"></iframe>


```{marginfigure}
You should turn a ZIP folder containing:  

- Your complete lab notebook as an .html and .Rmd file functioning with no errors.  
- Your data files.  
- Your completed assessment.
```

*An example notebook is available [here](https://alphonse.github.io/archive/chem370-f2020/notebooks/good-example/)*.

1. Use the GC-MS library (in ChemStation or OpenChrom) to get an initial idea of what compounds are present.
1. Move the data folder for each sample (ending in .D) to NEON. 
1. Transfer the data to your lab notebook folder and complete your analysis of the *2 most abundant compounds* in each sample.
1. Plot your chromatogram as *Relative Abundance (%)* vs. *Time (minutes)*.
1. Plot your mass spectra as *Relative Abundance (%)* vs. *m/z*.
1. Clearly identify the base peak, the molecular ion, and any important isotope peaks.  Also state how many rings and double bonds your molecule has, and whether it has an odd or even number of nitrogen atoms.  Provide a formula, common name, and molecular weight for each compound.  You must show your calculations for all of these items.  See the [mass spec guide](/archive/chem370-f2020/guides/mass-spec) for help with interpretation.  Does your interpretation match the library?

### Assessment {#vape-gc-assessment}

Answer the following questions by saving the result as the specified R variable in an R code block.  These questions are graded automatically.  If you do not name your variable correctly (including capitals) your questions will be marked incorrect.

Each question is worth 1 point.  You must get the question *exactly* right to get credit.

1. What is the mass-to-charge ratio of the base peak in your spectrum?  Call the variable `base`.

```{r question-gcms-1, purl = TRUE}
base <- 0     # put your answer between the '<-' and '#'
```

2. What is the mass-to-charge ratio of the molecular ion?  Call the variable `molion`.

```{r question-gcms-2, purl = TRUE}
molion <- 0     # put your answer between the '<-' and '#'
```

3. How many C atoms are present in your sample?  Call the variable `catoms`.

```{r question-gcms-3, purl = TRUE}
catoms <- 0     # put your answer between the '<-' and '#'
```

4. How many rings and/or double bonds are present in your sample? Call the variable `rdb`.

```{r question-gcms-4, purl = TRUE}
rdb <- 0     # put your answer between the '<-' and '#'
```

5. TRUE or FALSE: My molecule contains an EVEN number of N atoms.  Answer `TRUE` or `FALSE` in the variable `oddeven`.

```{r question-gcms-5, purl = TRUE}
oddeven <- FALSE     # put your answer between the '<-' and '#'
```

6. What is the monoisotopic molecular weight of your molecule? Call the variable `molwt`.

```{r question-gcms-6, purl = TRUE}
molwt <- 0     # put your answer between the '<-' and '#'
```

7. Propose a formula for your molecule.  Use the variables `c` for number carbon atoms, `h` for number of hydrogen, `o` for oxygen, and `n` for nitrogen.  Your sample will not contain atoms other than these for this lab activity.

```{r question-gcms-7, purl = TRUE}
c <- 0     # put your answer between the '<-' and '#'
h <- 0     # put your answer between the '<-' and '#'
o <- 0     # put your answer between the '<-' and '#'
n <- 0     # put your answer between the '<-' and '#'
```


\newpage

## Flame AA

Flame atomic absorption spectroscopy is a common method of determining the concentration of specific metals (cations) in a sample.  A solution is nebulized (aerosolized) into an acetylene flame where it is atomized (broken apart into its constituent atoms).  The atoms in the flame absorb characteristic wavelengths of light, unique to each element.  A light source specific to the analyte of interest (specifically, a hollow cathode lamp) shines through the flame, where the light gets absorbed by the sample.  The amount of absorption absorbed is linearly proportional to the concentration of the analyte in solution.  Although the instrument is relatively robust and inexpensive, it is limited to probing for a single analyte at a time.

The general procedure for this lab is to: (1) vape a sample and collect the aerosol, (2) prepare standards, blanks, and QCs, (3) run the standards, blank, and QC, and (4) run the samples.

For flame AA, the definitions of "passing" are:  

  1. **Blank:** Flat, smooth baseline showing only random noise and/or instrument drift no greater than a few percent over 10-15 minutes.  
  1. **Standard Curve:** A linear line with $R^2 \ge 0.9990$.  
  1. **QC:** A measured concentration within 5% ($\pm$ 5%) of the known concentration.  

This is a quantitative analysis.

### Safety {-}

- This experiment involves using an e-cigarette for an extended period of time.  The cigarette may get very hot over this time period; take care not to burn yourself.  
- Nicotine is toxic.  Wear gloves when working with the vape juice and handling the e-cigarette.  
- Do not breathe the e-liquid aerosol or let it into the room -- do the vaping step in a fume hood!  
- Never leave the flame AA unattended when the flame is on.  
- Although samples are over-the-counter products, you should avoid ingesting and/or contacting them with your skin.


### Standard Preparation

1. Prepare blanks from a mix of 90% ultrapure water/5% propylene glycol/5% glycerin (solution provided for you).
1. Prepare 5 zinc standards spanning the range of 1--10 ppm using the stock solution provided.  Dilute with 90% ultrapure water/5% propylene glycol/5% glycerin (solution provided for you) using volumetric glassware.
1. Prepare a QC from a different stock solution.  The concentration should fall somewhere near the middle of your standard curve.
1. Place your standards in 50 mL plastic centrifuge tubes for analysis.

### Vaping and Sample Preparation

***Note: you only need to do this procedure with ONE of your samples!***

##### **Pre-vape Sample** {-}

1. Place approximately 1 mL of e-liquid in a tared 15-mL centrifuge tube.
1. Obtain the mass of the liquid in the tube.
1. Fill the tube to 10 g with ultrapure water.
1. Filter the sample into a clean centrifuge tube using a 0.45 $\mu$m syringe filter.

##### **Post-vape Sample** {-}

1. Fill an e-cigarette about half full with your e-liquid.
1. Use the laboratory vaping apparatus (peristaltic pump and collection tube) to vape the sample into a tared 15-mL centrifuge tube.  Vape the sample in 5-second bursts followed by 30-second pauses to mimic an actual user.  Do this step in the fume hood!
1. Obtain the mass of the liquid in the tube.
1. Fill the tube to 10 g with ultrapure water.
1. Filter the sample into a clean centrifuge tube using a 0.45 $\mu$m syringe filter.

### Running Samples

1. Use the Flame AA SOP to prepare a flame AA *method* file containing your blanks and standards.  Use the `CHEM370_Vape` method as a template, but don't save over the template!
1. Prepare a *Sample Information File* containing your QC.
1. Run the blanks, standards, and QC.

    ```{marginfigure}
    **Checkpoint:**

    - A passing blank will show no signs of contamination.  
    - A passing standard curve will have $R^2 \ge 0.9990$.  
    - A passing QC will be within $\pm$ 5% of the known concentration.  

    If any of these is not passing you must repeat the analysis and/or remake your standards until they pass!
    ```

1. If everything passes, add your samples to your Sample Information File.
1. Run your newly-added samples.

*If you have extra sample, try screening it for other metals! (Ask how.)*

### Waste Disposal and Cleanup

1. Collect all wastes in the Flame AA waste container.
1. Place used (empty) centrifuge tubes in the trash.
1. Wipe up any spilled e-cigarette liquid with a paper towel (including liquid that's leaked onto the e-cigarette).
1. Remove the vaping machine tubing and rinse it out with ultrapure water.

### Data Processing {#faas-data-processing-vape}

The following video walks you through the data processing portion of this lab.

<iframe src="https://wcu.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=733fb481-8a20-4369-bf3c-ac67011c358e&autoplay=false&offerviewer=true&showtitle=true&showbrand=false&start=0&interactivity=all" height="405" width="720" frameBorder = "0" style="border: 0px solid #464646; display: block; margin: auto;" allowfullscreen allow="autoplay"></iframe>



```{marginfigure}
You should turn a ZIP folder containing:  

- Your complete lab notebook as an .html and .Rmd file functioning with no errors.  
- Your data files.  
- Your completed assessment.
```


1. Export your data to a CSV file using the **Data Manager** software package (shortcut on Desktop).  Use the `C370_template` design when prompted.
1. Use R to create and plot a calibration curve.
1. Determine the concentration of your samples and QCs.  Remember to factor in all dilutions.
1. Use the confidence intervals or a *t*-test (whichever you prefer) to determine whether the metal concentration is *significantly* different before and after vaping. (See Appendix 3, Comparing Means for more information on *t*-tests.)

### Assessment {#vape-faas-assessment}

Answer the following questions by saving the result as the specified R variable in an R code block.  These questions are graded automatically.  If you do not name your variable correctly (including capitals) your questions will be marked incorrect.

Each question is worth 1 point.  You must get the question *exactly* right to get credit.

1. What is $R^2$ value for your calibration curve?  Call the variable `r2`.

```{r question-faas-1, purl = TRUE}
r2 <- 0     # put your answer between the '<-' and '#'
```

2. What is sensitivity factor?  Call the variable `m`.

```{r question-faas-2, purl = TRUE}
m <- 0     # put your answer between the '<-' and '#'
```

3. What is mean measured analyte concentration for the QC in mg/L?  Call the variable `qc_conc`.

```{r question-faas-3, purl = TRUE}
qc_conc <- 0     # put your answer between the '<-' and '#'
```

4. What is the expected analyte concentration for the QC in mg/L?  Call the variable `qc_expected`.

```{r question-faas-4, purl = TRUE}
qc_expected <- 0     # put your answer between the '<-' and '#'
```

5. TRUE or FALSE: My QC passes at the 95% confidence level.  Save your `TRUE` or `FALSE` answer as `qc_passfail`.

```{r question-faas-5, purl = TRUE}
passfail <- FALSE     # put your answer between the '<-' and '#'
```

6. What is mean analyte concentration for sample 1 in mg/L?  Call the variable `s1_conc`.

```{r question-faas-6, purl = TRUE}
s1_conc <- 0     # put your answer between the '<-' and '#'
```

7. What is mean analyte concentration for sample 2 in mg/L?  Call the variable `s2_conc`.

```{r question-faas-7, purl = TRUE}
s2_conc <- 0     # put your answer between the '<-' and '#'
```

8. What is the 95% CI for sample 1 in mg/L?  Call the variable `s1_ci`.

```{r question-faas-8, purl = TRUE}
s1_ci <- 0     # put your answer between the '<-' and '#'
```

9. (Acetaminophen Only) What is the 95% CI for sample 2 in mg/L?  Call the variable `s2_ci`.

```{r question-faas-9, purl = TRUE}
s2_ci <- 0     # put your answer between the '<-' and '#'
```

10. TRUE or FALSE: The concentration of Sample 1 is significantly different from the labeled value.  Save your `TRUE` or `FALSE` answer as `s1_passfail`.

```{r question-faas-10, purl = TRUE}
s1_passfail <- FALSE     # put your answer between the '<-' and '#'
```

11. (Acetaminophen Only)  TRUE or FALSE: The concentration of Sample 2 is significantly different from the labeled value.  Save your `TRUE` or `FALSE` answer as `s2_passfail`.

```{r question-faas-11, purl = TRUE}
s2_passfail <- FALSE     # put your answer between the '<-' and '#'
```


\newpage

## FT-IR

FT-IR spectroscopy is used to deduce the bonds present in liquids, solids, and gases.  It is widely used in industry, research, and teaching for qualitative analysis, but can also provide very accurate quantitative information.  All molecules have a unique infrared spectrum, comparable to a person's fingerprint.  However, it is usually very difficult to interpret infrared spectra of mixtures.  Chemometrics and linear algebra can circumvent this issue by picking out minute differences between spectra using statistical methods.  During this lab, FT-IR will be used to match a sample of e-cigarette liquid to a reference spectrum using a metric called the hit quality index (HQI).  This methods provides a measure of how well two spectra match, and a HQI of 0.99 or better is required to indicate a contaminant-free sample.

This lab will use the Perkin Elmer Spectrum One FT-IR (Fourier-transform infrared spectrometer) to determine whether e-cigarette liquids are adulterated.  The Spectrum One is in NS 226.  Before lab, read *Standardization of Raman Spectra for Transfer of Spectral Libraries Across Different Instruments*.  Focus on the paragraph regarding hit-quality index calculation.

The general order for FT-IR sampling is: (1) collect a background (no sample present) and (2) collect sample spectra by placing a drop of sample on the instrument.  You will collect 3 spectra for each sample and 3 spectra of the reference sample.

### Safety {-}

- This lab will require the use of small amounts of [acetone](http://www.sciencelab.com/msds.php?msdsId=9927062); avoid contact with skin and eyes.
- Although samples are over-the-counter products, you should avoid ingesting and/or contacting them with your skin.

### Instrumentation

1.  Use the CHEM370 user on the FTIR.  Make sure you are collecting spectra as *Absorbance* vs. *Wavenumber*.
1.  Record a background spectrum on the FT-IR according to the FT-IR SOP.
1.  Place 1 drop the sample on the ATR window and acquire a spectrum. You do not need the anvil for liquid samples.
2.  Wet a Kimwipe with a small amount of acetone and clean the ATR cell with it.
3.  Repeat the above steps 2 more times, for a total of three spectra for your sample.
4.  Repeat the above steps for each sample, so you end up with three spectra of each sample.
4.  Make sure you leave the instrument and your work area clean, and log out of the computer.

### Waste Disposal and Cleanup

- Clean the IR with acetone and replace the cover.

  > Never spray acetone directly on the IR -- always spray it on a Kimwipe pointed away from the instrument.

- Wipe up any e-cigarette liquid on and around the IR with acetone.
- Make sure your work area and the instrument are cleaner than you found them.

### Data Processing {#ir-data-processing-vape}

The following video walks you through the data processing portion of this lab.

<iframe src="https://wcu.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=683a0991-3eb6-4e62-8bda-ac4600e1a364&autoplay=false&offerviewer=true&showtitle=true&showbrand=false&start=0&interactivity=all" height="405" width="720" frameBorder = "0" style="border: 0px solid #464646; display: block; margin: auto;" allowfullscreen allow="autoplay"></iframe>




```{marginfigure}
You should turn a ZIP folder containing:  

- Your complete lab notebook as an .html and .Rmd file functioning with no errors.  
- Your data files.  
- Your completed assessment.
```


- You will need to export you data as a CSV file.  The Spectrum One does this automatically.  Just copy the exported data from the "Spectrum One Data Folder" (there is a shortcut on the desktop) to your folder on NEON.
- See Appendix 2 for examples of the following steps:  
    - Baseline correct your spectra by subtracting the minimum value from each data point.  This should make the minimum value on your plot 0.
    - Normalize your spectra by converting from absorbance units to relative absorbance.  You do this by dividing the whole spectrum by the maximum absorbance value.  Your final plot should range from 0 to 1.  
    - Calculate the Hit Quality Index according to Eq. 2 in [Rodriguez et al (2011)](https://www.researchgate.net/publication/51602215_Standardization_of_Raman_spectra_for_transfer_of_spectral_libraries_across_different_instruments).  Be mindful of the difference between "x" (multiplication) and "·" (dot product).
    
### Assessment {#vape-ftir-assessment}

Answer the following questions by saving the result as the specified R variable in an R code block.  These questions are graded automatically.  If you do not name your variable correctly (including capitals) your questions will be marked incorrect.

Each question is worth 1 point.  You must get the question *exactly* right to get credit.

1. What is the unique ID (UID) value for Sample 1?  Call the variable `s1_uid`.

```{r question-ftir-1, purl = TRUE}
s1_uid <- 0     # put your answer between the '<-' and '#'
```

2. What is the unique ID (UID) value for Sample 2?  Call the variable `s2_uid`.

```{r question-ftir-2, purl = TRUE}
s2_uid <- 0     # put your answer between the '<-' and '#'
```

3. What is the hit quality index (HQI) value for Sample 1?  Call the variable `s1_hqi`.

```{r question-ftir-3, purl = TRUE}
s1_hqi <- 0     # put your answer between the '<-' and '#'
```

4. What is the hit quality index (HQI) value for Sample 2?  Call the variable `s2_hqi`.

```{r question-ftir-4, purl = TRUE}
s2_hqi <- 0     # put your answer between the '<-' and '#'
```

5. TRUE or FALSE: Sample 1 shows signs of contamination.  Save your `TRUE` or `FALSE` answer as `s1_passfail`.

```{r question-ftir-5, purl = TRUE}
s1_passfail <- FALSE     # put your answer between the '<-' and '#'
```

6. TRUE or FALSE: Sample 2 shows signs of contamination.  Save your `TRUE` or `FALSE` answer as `s2_passfail`.

```{r question-ftir-6, purl = TRUE}
s2_passfail <- TRUE     # put your answer between the '<-' and '#'
```

\newpage

## HPLC

This lab will use the Agilent 1220 HPLC (high performance liquid chromatograph) to determine the concentration of nicotine in vaping liquid.

HPLC is a separation method.  Like all chromatographic methods, it employs a mobile phase and a stationary phase.  An analyte dissolved in solution is injected onto a column (the stationary phase) and carried through the column by a solvent, called the eluent or mobile phase.  The column is packed with a substrate, which -- for reverse-phase HPLC as we'll perform here -- is a non-polar substance with a C<sub>18</sub> coating.  In the column, the sample partitions between the stationary and the mobile phase, and it's relative affinity for each determines how quickly it moves through the column.  In this way, HPLC allows us to separate analytes based on polarity.  In this lab, you'll use the HPLC to perform quantitative analysis.  This is possible because the area under each chromatographic peak is proportional to the concentration of each analyte.

The general order for HPLC sample preparation is: (1) Prepare standard and quality control samples (QCs), (2) run standards and a QC, (3) prepare the samples, and (4) run the samples.  You should obtain a clean blank, a passing standard curve, and a passing QC before preparing your samples.  If any of these fails there is no point in running your samples!

For HPLC, the definitions of "passing" are:  

  1. **Blank:** Flat, smooth baseline showing only random noise and/or instrument drift no greater than a few percent over 15 minutes.  
  1. **Standard Curve:** A linear line with $R^2 \ge 0.9900$.  
  1. **QC:** A measured concentration within 5% ($\pm$ 5%) of the known concentration.  

This is a quantitative analysis.

### Safety {-}

- Nicotine is toxic.  Wear gloves when working with the vape juice and preparing standards.  
- Although samples are over-the-counter products, you should avoid ingesting and/or contacting them with your skin.
- 75% water / 25% Acetonitrile / 0.1% formic acid is used as the HPLC mobile phase.  Avoid skin contact with mobile phase.

### Reference {-}

- [Davis et al (2015)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4892702/)

### Standard Preparation

1. Prepare blanks from ultrapure water.
1. Prepare nicotine standards ranging from 1--100 ppm using the stock solution provided.  Use ultrapure water as your solvent.
1. Use a separate stock solution to prepare a QC.  The nicotine concentration should fall toward the middle of your standard curve.
1. Place your standards in 2 mL HPLC autosampler vials for analysis (the ultrapure water is already filtered, so as long as you use fresh water you do not need to filter again!).

### Running Standards

1. Warm up the HPLC according to the HPLC SOP.  Use the `CHEM370_Vape.M` method.
1. Place your standards in the HPLC autosampler.
1. Prepare a sequence based on the autosampler locations of your samples (refer to HPLC SOP for more info).  Prepare your sequence in this order:
    1. Zero Standard/Rinse Blank (water)
    1. Low Standard
    1. Medium-low Standard
    1. Medium Standard
    1. Medium High Standard
    1. High Standard
    1. Rinse Blank
    1. QC Sample
1. After the instrument has warmed up (stable baseline), run your sequence.
1. Analyze your standard curve, blanks, and QC.

    ```{marginfigure}
    **Checkpoint:**

    - A passing blank will show no signs of contamination.  
    - A passing standard curve will have $R^2 \ge 0.9900$.  
    - A passing QC will be within $\pm$ 5% of the known concentration.  

    If any of these is not passing you must repeat the analysis and/or remake your standards until they pass!
    ```

### Sample Preparation

Repeat the following for each sample:

1. Tare an HPLC sample vial.
1. Place 2 drops of e-cigarette liquid in the vial and record the mass of liquid added.
1. Fill to top with approximately 2 mL ultrapure water.
1. Record the mass of the solvent you added.  You will need the masses to back calculate the undiluted nicotine concentration during data analysis.
1. Prepare enough samples to analyze all samples in triplicate.

    ```{marginfigure}
    Note that the ultrapure water and e-cigarette liquid are already filtered, so you do not need to filter them again during sample prep!
    ```

### Running Samples

1. Warm up the HPLC according the the HPLC SOP (skip if already warm).
1. Place your standards in the HPLC autosampler.
1. Prepare a sequence based on the autosampler locations of your samples (refer to HPLC SOP for more info).  Prepare your sequence in this order:
    1. Standards & QC (see *HPLC | Instrumentation: Standards* section)
    1. Up to 10 samples
    1. QC Sample (same as previous QC; always finish with QC)
1. After the instrument has warmed up, run your sequence.
1. Analyze your standards and samples (you may skip the standards again if you've run them in the past hour -- just start with a QC instead).
1. Use a *t*-test to determine if the mass of nicotine is *significantly* different from the expected mass. (See Appendix 3, Comparing Means for more information on *t*-tests.)

### Waste Disposal and Cleanup

1. Place all liquid wastes in the HPLC waste container for the lab.
1. Place all (cleaned) disposable glass vials in the broken class bin.

    > No chemicals should go in broken glass!

### Data Processing {#hplc-data-processing-vape}

The following video walks you through the data processing portion of this lab.

<iframe src="https://wcu.hosted.panopto.com/Panopto/Pages/Embed.aspx?id=bb4bb7a4-7260-4f97-9ec6-ac6300e32d3d&autoplay=false&offerviewer=true&showtitle=true&showbrand=false&start=0&interactivity=all" height="405" width="720" frameBorder = "0" style="border: 0px solid #464646; display: block; margin: auto;" allowfullscreen allow="autoplay"></iframe>

```{marginfigure}
You should turn a ZIP folder containing:  

- Your complete lab notebook as an .html and .Rmd files functioning with no errors.  
- Your data files.  
- Your completed assessment.
```

##### **Exporting & Plotting Representative Chromatograms** {-}

1. Move all your .D data folders onto NEON and then into your lab notebooks folder. 
1. Use ChemStation or OpenChrom to export CSVs of your chromatograms.

##### **Standard Curve and Concentrations** {-}

1. Use your standards to determine the retention time for nicotine.
1. Use the HPLC software (or R if you prefer) to integrate the nicotine peak for each sample.
1. Manually record the integration results in your lab notebook.
1. Calculate a standard curve using a linear model with a floating intercept.   Use units of ppm for your standard curve.
1. Calculate the concentration (in ppm) of nicotine in the QCs using your standard curve.
1. Calculate the concentration (in ppm) of nicotine in the samples using your standard curve.
1. Perform a *t*-test to determine in if the concentration of nicotine in your samples is *significantly* different than the expected amount.

##### **Limits of Detection and Quantitation** {-}

1. Calculate limits of detection and quantitation as you did in Lab 1.  Use the standard deviation of the entire blank as your standard deviation for the calculation. Report these in your technical report.

### Assessment {#vape-hplc-assessment}

Answer the following questions by saving the result as the specified R variable in an R code block.  These questions are graded automatically.  If you do not name your variable correctly (including capitals) your questions will be marked incorrect.

Each question is worth 1 point.  You must get the question *exactly* right to get credit.

1. What is $R^2$ value for your calibration curve?  Call the variable `r2`.

```{r question-hplc-1, purl = TRUE}
r2 <- 0     # put your answer between the '<-' and '#'
```

2. What is sensitivity factor?  Call the variable `m`.

```{r question-hplc-2, purl = TRUE}
m <- 0     # put your answer between the '<-' and '#'
```

3. What is mean measured analyte concentration for the QC in mg/L?  Call the variable `qc_conc`.

```{r question-hplc-3, purl = TRUE}
qc_conc <- 0     # put your answer between the '<-' and '#'
```

4. What is the expected analyte concentration for the QC in mg/L?  Call the variable `qc_expected`.

```{r question-hplc-4, purl = TRUE}
qc_expected <- 0     # put your answer between the '<-' and '#'
```

5. TRUE or FALSE: My QC passes at the 95% confidence level.  Save your `TRUE` or `FALSE` answer as `qc_passfail`.

```{r question-hplc-5, purl = TRUE}
passfail <- FALSE     # put your answer between the '<-' and '#'
```

6. What is mean analyte concentration for sample 1 in mg/L?  Call the variable `s1_conc`.

```{r question-hplc-6, purl = TRUE}
s1_conc <- 0     # put your answer between the '<-' and '#'
```

7. What is mean analyte concentration for sample 2 in mg/L?  Call the variable `s2_conc`.

```{r question-hplc-7, purl = TRUE}
s2_conc <- 0     # put your answer between the '<-' and '#'
```

8. What is the 95% CI for sample 1 in mg/L?  Call the variable `s1_ci`.

```{r question-hplc-8, purl = TRUE}
s1_ci <- 0     # put your answer between the '<-' and '#'
```

9. What is the 95% CI for sample 2 in mg/L?  Call the variable `s2_ci`.

```{r question-hplc-9, purl = TRUE}
s2_ci <- 0     # put your answer between the '<-' and '#'
```

10. TRUE or FALSE: The concentration of Sample 1 is significantly different from the labeled value.  Save your `TRUE` or `FALSE` answer as `s1_passfail`.

```{r question-hplc-10, purl = TRUE}
s1_passfail <- FALSE     # put your answer between the '<-' and '#'
```

11. TRUE or FALSE: The concentration of Sample 2 is significantly different from the labeled value.  Save your `TRUE` or `FALSE` answer as `s2_passfail`.

```{r question-hplc-11, purl = TRUE}
s2_passfail <- FALSE     # put your answer between the '<-' and '#'
```
